---
title: "Fruit and Vegetable Consumption in Canada — R"
subtitle: "A simple analysis using R"
author: "Patrick Pascal"
date: "2026-09-28"
categories: [R, Nutrition, Canada]
engine: knitr
execute:
  echo: true
---

## Research question

How did the percentage of Canadians reporting fruit and vegetable
consumption five or more times per day change between 2015 and 2021?

This post repeats the same analysis from my Python post using R.

## Data source

The data come from Statistics Canada's
[Canadian Indicator Framework, indicator 3.1.1](https://sdgcif-data-canada-oddcic-donnee.github.io/3-1-1/).

The indicator measures reported eating frequency, not portion sizes
or the total amount consumed.

The data are available under the
[Statistics Canada Open Licence](https://www.statcan.gc.ca/en/terms-conditions/open-licence).

Adapted from Statistics Canada, Canadian Indicator Framework,
indicator 3.1.1, reference years 2015–2021. This does not constitute
an endorsement by Statistics Canada of this product.

## R setup

I load the packages needed to read, prepare, display, and plot the data.

```{r}
#| label: setup-r
#| message: false

library(readr)
library(dplyr)
library(ggplot2)
library(knitr)

.libPaths()
```

## Load and inspect the data

I read the same CSV used in the Python analysis and display the
available observations.

```{r}
#| label: load-nutrition-data-r

nutrition_raw <- read_csv(
  "../../data/fruit_vegetable_canada.csv",
  show_col_types = FALSE
)

knitr::kable(nutrition_raw)
```

## Prepare the data

I keep the available observations from 2015 through 2021,
select the year and percentage columns, and sort the data by year.

```{r}
#| label: prepare-nutrition-data-r

nutrition <- nutrition_raw |>
  filter(Year >= 2015, Year <= 2021) |>
  select(Year, Percentage = Value) |>
  arrange(Year)

knitr::kable(nutrition)
```

The prepared data contain five observations. There are no observations
for 2018 or 2019 in the downloaded dataset, so those years remain missing.

## Calculate the change

I compare the first and last percentages in the prepared data.

```{r}
#| label: calculate-nutrition-change-r

first_percentage <- nutrition$Percentage[1]
last_percentage <- nutrition$Percentage[nrow(nutrition)]

change_pp <- last_percentage - first_percentage

cat(sprintf("2015: %.1f%%\n", first_percentage))
cat(sprintf("2021: %.1f%%\n", last_percentage))
cat(sprintf("Change: %.1f percentage points\n", change_pp))
```

The reported percentage decreased from 31.5% in 2015 to 21.8% in 2021,
a decrease of 9.7 percentage points.

This calculation describes the difference between the published
estimates. It does not explain why the estimates changed.

## Visualize the trend

Each point represents an available published observation.
The connecting line helps show the pattern but does not estimate
the missing years.

```{r}
#| label: fig-nutrition-r
#| fig-cap: "Percentage reporting fruit and vegetable consumption five or more times per day, Canada (excluding territories), available years from 2015 to 2021. Source: Statistics Canada."
#| fig-alt: "Five observations declining from 31.5 percent in 2015 to 21.8 percent in 2021, with no observations for 2018 or 2019."

ggplot(nutrition, aes(x = Year, y = Percentage)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 2015:2021) +
  scale_y_continuous(limits = c(0, 40)) +
  labs(
    title = "Fruit and vegetable consumption in Canada",
    x = "Year",
    y = "Reported percentage (%)"
  ) +
  theme_minimal()
```

## Conclusion

The published percentage reporting fruit and vegetable consumption
five or more times per day decreased from 31.5% in 2015 to 21.8%
in 2021, a decline of 9.7 percentage points.

The percentage was lower at each successive available observation.
However, the dataset has no observations for 2018 or 2019, so it does
not show what happened in every individual year.

This is a descriptive comparison of published estimates. The indicator
measures reported eating frequency rather than overall diet quality,
and this analysis does not establish why the estimates changed.